CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-07-24 script/de/chapters/numbertheory.tex
no message

Dieser Anhang beschreibt, wie man die Basisdarstellung einer nat"urlichen Zahl umwandelt
und wie man mit Hilfe der Logarithmusfunktion die Ziffernl"ange einer beliebigen 
nat"urlichen Zahl absch"atzen kann.

@@

Wenn man diese Ziffern $a_na_{n-1} \cdots a_1a_0$ direkt hintereninaderschreibt und die Gewichte $b^i$ wegl"asst, 
so erh"alt man die gewohnte Schreibweise f"ur Zahlen.

@@

\subsubsection*{L"ange der Zifferndarstellung}
F"ur eine nat"urliche Zahl $z$ kann die L"ange der Zahl in $b$-adischer Darstellung wie folgt
bestimmt werden. Dazu geht man von der Absch"atzung $b^{n+1} > z \ge b^n$, wobei $n+1$ die gesuchte Stellenzahl ist, aus.
Nach dem Logarithmieren zur Basis $b$\footnote{% 
Nach den Logarithmengesetzen gilt f"ur die Basen $b$ und $b'$ die Gleichung $\log_b z = \log_{b'} z / \log_{b'} (b)$. 
Es ist also einfach, ausgehend von Logarithmentafeln f"ur z.B. die Basis $b' = 10$ den Logarithmus zur Basis $b = 2$ zu berechnen.
} bestimmt man die Ungleichung $n+1 > log_b z \ge n$. 
Es folgt, dass $n = \lfloor \log_b z \rfloor$\footnote{%
Die Funktion $\lfloor x \rfloor$ bestimmt die n"achste ganzen Zahl, kleiner $x$ (wenn $x \ge 0$, dann werden die Nachkommastellen 
der Zahl einfach abgeschnitten). Man sagt auch {\em untere Gau"sklammer von} $x$.
}. Man bezeichnet mit $l_b(z)$ die Zahl der ben"otigten Ziffern zur Darstellung
der Zahl $z$ in $b$-adischer Schreibweise. Es gilt die Gleichung $$l_b(z) = \lfloor \log_b z \rfloor +1. $$

@@

\vskip +10 pt
{\bf Beispiel 1}:
Die Zahl $z = 234$ in Dezimalschreibweise wird umgewandelt in 
Hexadezimalschreibweise. $a_0 = 234 {\rm ~mod~} 16 = 10 = A$, $234 / 16 = 15 = E$, 
$a_1 = 14 {\rm ~mod~} 16 = E$ also $EA$.

@@

Das schnelle Potenzieren modulo $m$ geh"ort zu den Grundfunktionen von Mathematica und 
Pari-GP. Sie k"onnen mit Hilfe dieser Programme nat"urlich auch die Idee der
Square-and-Multiply-Methode nachvollziehen. In Mathematica bestimmen Sie die Einzelexponentiation des Beispiels auf 
Seite \pageref{SrcArith2} mit
\begin{verbatim}
Mod[{87^43, 87^2, 87^4, 87^8, 87^16, 87^32}, 103] = {85, 50, 28, 63, 55, 38}.Die L"ange der Hexadezimalschreibweise bestimmt man anhand von $z$ durch die Berechnung
$$l_{16}(z) = \lfloor \log_{16}(z) \rfloor + 1 = \lfloor \ln (z) / \ln(16) \rfloor + 1 = \lfloor 1,96... \rfloor + 1 = 1 + 1 = 2.$$

\end{verbatim}
und in Pari-GP lautet die Syntax:
\begin{verbatim}
Mod([87^43,87^2,87^4,87^8,87^16,87^32],103)
\end{verbatim}

>>>>

Dieser Anhang beschreibt, wie man die Basisdarstellung einer nat"urlichen Zahl umwandelt
und wie man mit Hilfe der Logarithmusfunktion die Ziffernl"ange jeder nat"urlichen Zahl 
zu einer beliebigen Basis absch"atzen kann.

@@

Wenn man diese Ziffern $a_na_{n-1} \cdots a_1a_0$ direkt hintereinanderschreibt und die Gewichte $b^i$ wegl"asst, 
so erh"alt man die gewohnte Schreibweise f"ur Zahlen.

@@

\subsubsection*{L"ange der Zifferndarstellung}
F"ur eine nat"urliche Zahl $z$ kann die L"ange der Zahl in $b$-adischer Darstellung wie folgt
bestimmt werden. Dazu geht man von der Absch"atzung $b^{n+1} > z \ge b^n$, wobei $n+1$ die gesuchte Stellenzahl ist, aus.
Nach dem Logarithmieren zur Basis $b$\footnote{% 
Nach den Logarithmengesetzen gilt f"ur die Basen $b$ und $b'$ die Gleichung $\log_b z = \log_{b'} z / \log_{b'} (b)$. 
Es ist also einfach, ausgehend von Logarithmentafeln f"ur z.B. die Basis $b' = 10$ den Logarithmus zur Basis $b = 2$ zu berechnen.
} bestimmt man die Ungleichung $n+1 > log_b z \ge n$. 
Es folgt, dass $n = \lfloor \log_b z \rfloor$\footnote{%
Die Funktion $\lfloor x \rfloor$ bestimmt die n"achste ganze Zahl kleiner $x$ (wenn $x \ge 0$, dann werden die Nachkommastellen 
der Zahl einfach abgeschnitten). Man sagt auch {\em untere Gau"sklammer von} $x$.
}. Man bezeichnet mit $l_b(z)$ die Zahl der ben"otigten Ziffern zur Darstellung
der Zahl $z$ in $b$-adischer Schreibweise. Es gilt die Gleichung $$l_b(z) = \lfloor \log_b z \rfloor +1. $$

@@

\vskip +10 pt
{\bf Beispiel 1}:
Die Zahl $z = 234$ in Dezimalschreibweise wird umgewandelt in 
Hexadezimalschreibweise: $a_0 = 234 {\rm ~mod~} 16 = 10 = A$, $234 / 16 = 15 = E$, 
$a_1 = 14 {\rm ~mod~} 16 = E$ also $EA$.

@@

Das schnelle Potenzieren modulo $m$ geh"ort zu den Grundfunktionen von Mathematica und 
Pari-GP. Sie k"onnen mit Hilfe dieser Programme nat"urlich auch die Idee der
Square-and-Multiply-Methode nachvollziehen. In Mathematica bestimmen Sie die Einzelexponentiation des Beispiels auf 
Seite \pageref{SrcArith2} mit
\begin{verbatim}
Mod[{87^43, 87^2, 87^4, 87^8, 87^16, 87^32}, 103] = {85, 50, 28, 63, 55, 38}
\end{verbatim}
und in Pari-GP lautet die Syntax:
\begin{verbatim}
Mod([87^43,87^2,87^4,87^8,87^16,87^32],103)
\end{verbatim}

### 2003-07-24 script/en/chapters/numbertheory.tex
no message

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_A}{} \label{Appendix_A}
\subsection*{Appendix A: the greatest common divisor (gcd) of whole numbers and the two
Algorithms of Euclid} 
\addcontentsline{toc}{subsection}{Appendix A: gcd of whole numbers and the two
Algorithms of Euclid}\index{Euclidean algorithm!extended}\index{Gcd}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\subsection*{Appendix D: Base representation of numbers, estimation of length of digits}
\addcontentsline{toc}{subsection}{Appendix D: Base representation and base transformation of numbers, estiimation of length of digits}

@@

We represent numbers usually in the decimal system (base 10). Computers
are working with the binary representation of numbers --- only for the display numbers are 
represented in decimal or sometimes hexadecimal (base 16).

This appendix describes how to generate arbitrary base representations of a positive integers and how 
to determine the number of required digits via the logarithm function.

\subsubsection*{$b$-adic sum representation of positive integers}
Given base $b$, each positive integer $z$ can be represented as a $b$-adic sum 
$$ z = a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0, $$
where $a_i \in \{0, 1, \dots, b-1\}, \; i=0,1,\dots, n$ are called {\em digits}.
For this sum, it follows that for arbitrary digits $a_0, a_1, \dots, a_n$ we have 
$b^{n+1} > a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0$. On the other hand there exist
$a_0, a_1, \dots, a_n$ (namely $a_i = b-1$ for $i=0, \dots, n$) following that  
$b^{n+1} -1 \le a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0$
(using these inequalities it can be shown that each positve integer can be represented by a $b$-adic sum).

@@

{\bf Examples}: \\
Base $b = 10$: $10278 =  1\cdot10^4 +  0\cdot 10^3 + 2\cdot 10^2 + 7\cdot 10^1 + 8$ und \\
Base $b = 16$: $FE70A = 15\cdot16^4 + 14\cdot 16^3 + 7\cdot 16^2 + 0\cdot 16^1 + 10$.

@@

For a positive integer $z$ the length of the $b$-adic representation can be determined via the 
following steps. Starting from the inequality $b^{n+1} > z \ge b^n$ 
we have --- after applying the logarithm function on basis $b$\footnote{%
Applying the logarithm formula on base $b$ and $b'$ we have $\log_b z = \log_{b'} z / \log_{b'} (b)$. 
It is therefore easy using e.g. logarithm tables for the base $b' = 10$ to compute the logarithm of base $b = 2$.
} --- $n+1 > log_b z \ge n$. Therefore we have $n = \lfloor log_b z \rfloor$\footnote{%
The function $\lfloor x \rfloor$ determines the next integer smaller than $x$ (in case $x \ge 0$ the digits after the decimal point are truncated). 
}. We call $l_b(z)$ {\em the number of required digits to represent the number $z$ on the base $b$}. We have
$$l_b(z) := \lfloor log_b z \rfloor +1.$$

\vskip +10 pt
{\bf Example 1}:
We compute for the decimal number $z = 234$ the hexadecimal representation (number base $b = 16$)
$$l_{16}(z) = \lfloor \log_{16}(z) \rfloor + 1 = \lfloor \ln (z) / \ln(16) \rfloor + 1 = \lfloor 1.96... \rfloor + 1 = 1 + 1 = 2.$$

\vskip +10 pt
{\bf Example 2}:
We compute for the decimal number $z = 234$ the binary representation (number base $b = 2$)
$$l_{2}(z) = \lfloor \log_{2}(z) \rfloor + 1 = \lfloor \ln (z) / \ln(2) \rfloor + 1 = \lfloor 7.87... \rfloor + 1 = 7 + 1 = 8.$$

@@

\vskip +10 pt
{\bf Example 2}:
The binary number $z = 1000100101110101$ is computed into the decimal representation via the following steps:

>>>>

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_A}{} \label{Appendix_A}
\subsection*{Appendix A: the greatest common divisor (gcd) of whole numbers and the two
Algorithms of Euclid} 
\addcontentsline{toc}{subsection}{Appendix A: gcd of whole numbers and the two
algorithms of Euclid}\index{Euclidean algorithm!extended}\index{Gcd}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\subsection*{Appendix D: Base representation of numbers, estimation of length of digits}
\addcontentsline{toc}{subsection}{Appendix D: Base representation and base transformation of numbers, estimation of length of digits}

@@

We represent numbers usually in the decimal system (base 10). Computers
are working with the binary representation of numbers --- only for the display numbers are 
represented in decimal or sometimes hexadecimal (base 16) form.

This appendix describes how to generate arbitrary base representations of anay positive integer and how 
to determine the number of required digits via the logarithm function.

\subsubsection*{$b$-adic sum representation of positive integers}
Given base $b$, each positive integer $z$ can be represented as a $b$-adic sum 
$$ z = a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0, $$
where $a_i \in \{0, 1, \dots, b-1\}, \; i=0,1,\dots, n$ are called {\em digits}.
For this sum, it follows that for arbitrary digits $a_0, a_1, \dots, a_n$ we have 
$b^{n+1} > a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0$. On the other hand there exist
$a_0, a_1, \dots, a_n$~(namely $a_i = b-1$ for $i=0, \dots, n$) following that  
$b^{n+1} -1 \le a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0$
(using these inequalities it can be shown that each positve integer can be represented by a $b$-adic sum).

@@

{\bf Examples}: \\
Base $b = 10$: $10278 =  1\cdot10^4 +  0\cdot 10^3 + 2\cdot 10^2 + 7\cdot 10^1 + 8$ and \\
Base $b = 16$: $FE70A = 15\cdot16^4 + 14\cdot 16^3 + 7\cdot 16^2 + 0\cdot 16^1 + 10$.

@@

For a positive integer $z$ the length of the $b$-adic representation can be determined via the 
following steps. Starting from the inequality $b^{n+1} > z \ge b^n$ 
we have --- after applying the logarithm function on basis $b$\footnote{%
Applying the logarithm formula on base $b$ and $b'$ we have $\log_b z = \log_{b'} z / \log_{b'} (b)$. 
It is therefore easy using e.g. logarithm tables for the base $b' = 10$ to compute the logarithm of base $b = 2$.
}:~$n+1 > log_b z \ge n$. Therefore we have $n = \lfloor log_b z \rfloor$\footnote{%
The function $\lfloor x \rfloor$ determines the next integer smaller than $x$ (in case $x \ge 0$ the digits after the decimal point are truncated). 
}. We call $l_b(z)$ {\em the number of required digits to represent the number $z$ on the base $b$}. We have
$$l_b(z) := \lfloor log_b z \rfloor +1.$$

\vskip +10 pt
{\bf Example 1}:
We compute for the decimal number $z = 234$ (EA in hex) the hexadecimal representation (number base $b = 16$)
$$l_{16}(z) = \lfloor \log_{16}(z) \rfloor + 1 = \lfloor \ln (z) / \ln(16) \rfloor + 1 = \lfloor 1.96... \rfloor + 1 = 1 + 1 = 2.$$

\vskip +10 pt
{\bf Example 2}:
We compute for the decimal number $z = 234$ (11101010 in binary) the binary representation (number base $b = 2$)
$$l_{2}(z) = \lfloor \log_{2}(z) \rfloor + 1 = \lfloor \ln (z) / \ln(2) \rfloor + 1 = \lfloor 7.87... \rfloor + 1 = 7 + 1 = 8.$$

@@

\vskip +10 pt
{\bf Example 2}:
The binary number $z = 1000100101110101$ is transformed into the decimal representation via the following steps:
